---
title: "Assignment 3 - Causal inference"
author: "RF"
date: "2/5/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment 3 - Exploring causal inference issues

In this assignment we explore some issues related to multiple regressions (regressions with more than one predictor), and inferred (causal) relations between variables. N.B. the data is simulated (to make sure I know the actual mechanism generating it), but it's based on a real study. So bear with a longish introduction to get into the details of what we are doing and why it is important.

### Altercentric intrusion in schizophrenia

People with schizophrenia often report altered control and distinction of self-other representations: intrusive thoughts, hearing of voices, delusions of mind reading, paranoia, etc (a substantial portion of the psychotic symptoms experienced in schizophrenia). These have been variously attributed to hypermentalizing (over attribution of mental states to others), social impairment (over preoccupation with own thought processes), hyper socialization (inability to inhibit information from others), etc.

The current study investigates 1) whether schizophrenia is indeed related to altered control and distinction of self-other representations, in particular altercentric intrusions (inability to inhibit social information), and 2) whether these are related to the relevant psychotic symptoms. N.B. the actual study also investigates egocentric intrusion, do check the papers below if interested.

The task is a slightly modified version of this: https://www.ncbi.nlm.nih.gov/pubmed/20731512 You look at a picture with some dots visible to you, as well as with a different person with a different set of dots visible to them. The number of dots you see and that the other sees can be the same (congruent condition) or not (incongruent condition). You are tasked to indicate whether a given number (e.g. 3) matches the number of dots you see (and the dots visible to the other person are irrelevant to the task).


The tasks investigates altercentric intrusion: will your reaction time change according to whether the other person is seeing the same amount of dots as you, or not? The idea is that if you correctly inhibit social information, your reaction time should not change, as the information about the other person is not relevant. On the contrary, if you nevertheless use task irrelevant social information, you'll be slower at indicating whether 3 is the right number of dots when the other person sees a different amount of dots than you (conflicting information).
The bigger the difference between RTs in the congruent and incongruent condition the bigger the altercentric intrusion effect.

For each participant you have 6 variables: 1) ID, 2) AltercentricIntrusion (continuous score), 3) Diagnosis (schizophrenia vs. control), 4) VoiceHearing (severity of voice hearing symptoms, continuous score of the severity of the symptom as measured by a clinician), 5) MindReading (severity of delusions of mind reading, continuous score of the severity of the symptom as measured by a clinician); 6) Apathy (severity of lack of motivation in taking care of oneself, from washing to showing up at work, continuous score of the severity of the symptom as measured by a clinician).

The research questions you have to answer are the following:

## First part

Q1.1) Does schizophrenia involved altercentric intrusion? Define model and priors. Test the implications of your priors (prior predictive checks) and if needed adjust them. Run the model. Test the quality of the fitted model (posterior predictive checks). Assess the evidence in favor of an increased altercentric intrusion in schizophrenia. Report the model and the results, including plots.

```{r}
pacman::p_load(tidyverse, brms, PerformanceAnalytics, parallel, vcov, GridExtra)

set.seed(123)

cores<- detectCores()#for parallel processing
# Prepare the data

d <- read_csv("Ass3.csv")
summary(d)

d$Diagnosis <- plyr::revalue(as.character(d$Diagnosis), 
                             c("0"="Controls", "1"="Schizophrenia"))

d <- d %>%
  mutate(
    ID = as.factor(ID),
    Diagnosis = as.factor(Diagnosis)
  )


summary(d)
# Define the formula
# Define the formula

AltercentricDiagnosis_f0 <- bf(
  AltercentricIntrusion ~ 1 + Diagnosis
)

AltercentricDiagnosis_f <- bf(
  AltercentricIntrusion ~ 0 + Diagnosis
)


# Design the priors

get_prior(AltercentricDiagnosis_f0, family = gaussian, d)
get_prior(AltercentricDiagnosis_f, family = gaussian, d)

priorDiagnosis <- c(
  prior(normal(4, 1), class = b), # mean and then 2SD - inspect the data
  prior(normal(1, 2), class = sigma)
) 

# Test the priors

AltercentricDiagnosis_PriorCheck_m <- brm(
  formula = AltercentricDiagnosis_f,
  data = d,
  family = gaussian,
  prior = priorDiagnosis,
  sample_prior = "only", #meaning we want to sample the prior
  cores = cores,
  file = "AltercentricDiagnosis_PriorCheck_m"
)
#
pp_check(AltercentricDiagnosis_PriorCheck_m, nsamples = 100)+ggtitle("AI~Diagnosis PriorCheck")

## Fitting the model
AltercentricDiagnosis_m <- brm(
  formula = AltercentricDiagnosis_f,
  data = d,
  family = gaussian,
  prior = priorDiagnosis,
  sample_prior = T,
  cores = cores,
  file = "AltercentricDiagnosis_m"
)

#plot model
plot(AltercentricDiagnosis_m)

# Posterior predictive check
pp_check(AltercentricDiagnosis_m, nsamples = 100)+ggtitle("AI~Diagnosis Posterior predictive check")

## Check the model for warnings
AltercentricDiagnosis_m

# Hypothesis testing + updating check
plot(hypothesis(AltercentricDiagnosis_m,
           "DiagnosisSchizophrenia > DiagnosisControls"))[[1]]+ggtitle("Is AI higher in Schizophrenia patients?")

hypothesis(AltercentricDiagnosis_m,
           "DiagnosisSchizophrenia > DiagnosisControls")

plot(conditional_effects(AltercentricDiagnosis_m), points=T,plot=F)[[1]]+ggtitle("Estimates of AI ~ Diagnosis")

#evidence ratio - the posterior is a bunch of samples and we want to count how many of them are in/compatible with our hypothesis and we count them and gives us the ratio (21 min)
```

The model indicates a credible difference in altercentric intrusion in the two groups supporting our hypothesis (b = 0.36, CIs = 0.16, 0.57, ER = 1332). Controls showed on average an altercentric intrusion effect of 3.86 (CIs 3.74, 3.98), and schizophrenia of 4.22 (CIs = 4.01, 4.43).
[Add plot of the effects]

# SI
The model had no divergences, a Rhat of 1, and Effective Sample Sizes above 2000 for both Bulk and Tail.
[Add prior and posterior checks plots; add updating check plot]

Q1.2) Is altercentric intrusion related to specific symptoms *in the patients*? Identify which of the symptoms could be relevant. Should you include more than one symptom? Build models, priors, predictive checks. Assess the evidence and report models and results, including plots. Discuss whether the results make sense.

```{r}
## Isolate disorder group
d <-  d %>% 
  mutate(
    AltercentricIntrusion = scale(AltercentricIntrusion),
    VoiceHearing = scale(VoiceHearing),
    MindReading = scale(MindReading),
    Apathy = scale(Apathy)
  )

schizoData <- d %>% 
  filter(Diagnosis == "Schizophrenia")


## define formula and priors ##

(schizoSummary<- psych::describe(schizoData))

# voice hearing
AI_VoiceHearing_f1 <- bf(
  AltercentricIntrusion ~ 1 + VoiceHearing)


# MindReading
AI_MindReading_f1 <- bf(
  AltercentricIntrusion ~ 1 + MindReading)


# Apathy
AI_Apathy_f1 <- bf(
  AltercentricIntrusion ~ 1 + Apathy)

# VH + MR
AI_VH_MR_f1 <- bf(
  AltercentricIntrusion ~ 1 + VoiceHearing + MindReading)

# ALL
AI_VH_MR_A_f1 <- bf(
  AltercentricIntrusion ~ 1 + Apathy + MindReading + VoiceHearing)


## Design the priors ##

# getting quick summary plus SD


# define prior

priorR2 <- c(
  prior(normal(0, 1), class = Intercept), # mean and then 2SD - inspect the data
  prior(normal(1, 2), class = sigma),
  prior(normal(0, .3), class = b) # sigma = the average error we expect
)

## a function for automatizaying prior check

check_prior <- function(formula, data, prior, filename){
  model <- brm(
  formula = formula,
  data = data,
  family = gaussian,
  prior = prior,
  sample_prior = "only",
  control = list( adapt_delta = 0.95),
  core = cores,
  file = filename)
  
  p_check <- pp_check(model, nsamples = 100)
  return(p_check)
}

## prior checking

checkprior_VH <- check_prior(AI_VoiceHearing_f1, schizoData, priorR2, "checkprior_VH")
checkprior_VH+ggtitle("AI~VoiceHearing Prior check")

checkprior_MR <- check_prior(AI_MindReading_f1, schizoData, priorR2, "checkprior_MR")
checkprior_MR+ggtitle("AI~MindReading Prior check")

checkprior_Apathy <- check_prior(AI_Apathy_f1, schizoData, priorR2, "checkprior_Apathy")
checkprior_Apathy+ggtitle("AI~Apathy Prior check")

checkprior_VH_MR <- check_prior(AI_VH_MR_f1, schizoData, priorR2, "checkprior_VH_MR")
checkprior_VH_MR+ggtitle("AI~VoiceHearing+MindReading Prior check")

checkprior_VC_MR_A <- check_prior(AI_VH_MR_A_f1, schizoData, priorR2, "checkprior_VC_MR_A")
checkprior_VC_MR_A+ggtitle("AI~VoiceHearing+MindReading+Apathy Prior check")

```


## fitting
```{r}

### voice hearing
VoiceHearing1 <- brm(
    formula = AI_VoiceHearing_f1,
    data = schizoData,
    family = gaussian,
    prior = priorR2,
    sample_prior = T,
    refresh = 0,
    cores = cores,
    file = "VoiceHearing1"
    
  )

#plot model

plot(VoiceHearing1)
  
# a plot for a posterior check
VH_posch <- pp_check(VoiceHearing1, nsamples = 100)
VH_posch+ggtitle("Ai~VoiceHearing Posterior Check")
  
# test and plot your hypothesis
VH_hyp <- hypothesis(VoiceHearing1, "VoiceHearing > 0")
VH_hyp

plot(conditional_effects(VoiceHearing1), points = T)[[1]]+ggtitle("Ai~VoiceHearing")
  

## mind reading ##
MindReading1 <- brm(
    formula = AI_MindReading_f1,
    data = schizoData,
    family = gaussian,
    prior = priorR2,
    sample_prior = T,
    refresh = 0,
    cores = cores,
    file = "MindReading1"
    )

MindReading1

#plot model

plot(MindReading1)
# a plot for a posterior check
MR_posch <- pp_check(MindReading1, nsamples = 100)
MR_posch+ggtitle("Ai~MindReading Posterior Check")
  
  
# test and plot your hypothesis
MR_hyp <- hypothesis(MindReading1, "MindReading > 0")
MR_hyp


plot(MR_hyp)
plot(conditional_effects(MindReading1), points = T)[[1]]+ggtitle("Ai~MindReading")
  

## apathy ##
Apathy1 <- brm(
    formula = AI_Apathy_f1,
    data = schizoData,
    family = gaussian,
    prior = priorR2,
    sample_prior = T,
    refresh = 0,
    cores = cores,
    file = "Apathy1"
    )

Apathy1

#plot model

plot(Apathy1)

# a plot for a posterior check
A_posch <- pp_check(Apathy1, nsamples = 100)
A_posch+ggtitle("Ai~Apathy Posterior Check")
  
  
# test and plot your hypothesis
A_hyp <- hypothesis(Apathy1, "Apathy < 0")
A_hyp

plot(A_hyp)  
plot(conditional_effects(Apathy1), points = T)[[1]]+ggtitle("Ai~Apathy")
```


```{r}
## voice hearing, mind reading

AI_VH_MR_m1 <- brm(
  formula = AI_VH_MR_f1,
  data = schizoData,
  family = gaussian,
  prior = priorR2,
  sample_prior = T,
  cores = cores,
  file = "AI_VH_MR_m1"
)
AI_VH_MR_m1

#plot model

plot(AI_VH_MR_m1)

# Posterior predictive check
pp_check(AI_VH_MR_m1, nsamples = 100)+ggtitle("AI~VoiceHearing+MindReading Posterior Predictive Check")

# Hypothesis testing + updating check

hypothes_VHMR_VH <- hypothesis(AI_VH_MR_m1,
           "VoiceHearing > 0")

hypothes_VHMR_MR <- hypothesis(AI_VH_MR_m1,
           "MindReading > 0")

hypothes_VHMR_VH
hypothes_VHMR_MR

plot(hypothes_VHMR_VH,plot=F)[[1]]+ggtitle("AI~VoiceHearing+MindReading")
plot(hypothes_VHMR_MR,plot=F)[[1]]+ggtitle("AI~VoiceHearing+MindReading")


## conditional effects
plot(conditional_effects(AI_VH_MR_m1), points = T)


## Voice hearing, mind reading, apathy
AI_VH_MR_A_m1 <- brm(
  formula = AI_VH_MR_A_f1,
  data = schizoData,
  family = gaussian,
  prior = priorR2,
  sample_prior = T,
  cores = cores,
  file = "AI_VH_MR_A_m1"
)


AI_VH_MR_A_m1

#plot model

plot(AI_VH_MR_A_m1)


# Posterior predictive check
pp_check(AI_VH_MR_A_m1, nsamples = 100)+ggtitle("AI~VoiceHearing+MindReading+Apathy Posterior Predictive Check")

# Hypothesis testing + updating check

hypothes3_VH <- hypothesis(AI_VH_MR_A_m1,
           "VoiceHearing > 0")

hypothes3_MR <- hypothesis(AI_VH_MR_A_m1,
           "MindReading > 0")

hypothes3_Apathy <- hypothesis(AI_VH_MR_A_m1,
           "Apathy < 0")


hypothes3_VH
hypothes3_MR
hypothes3_Apathy


plot(hypothes3_VH,plot=F)[[1]]+ggtitle("AI~VoiceHearing+MindReading+Apathy")
plot(hypothes3_MR,plot=F)[[1]]+ggtitle("AI~VoiceHearing+MindReading+Apathy")
plot(hypothes3_Apathy,plot=F)[[1]]+ggtitle("AI~VoiceHearing+MindReading+Apathy")


## conditional effects
plot(conditional_effects(AI_VH_MR_A_m1), points = T)
```

## Model comparison
```{r}
## adding criterion loo
VoiceHearing1 <- add_criterion(VoiceHearing1, criterion = "loo")
MindReading1 <- add_criterion(MindReading1, criterion = "loo")
Apathy1 <- add_criterion(Apathy1, criterion = "loo")
AI_VH_MR_m1 <- add_criterion(AI_VH_MR_m1, criterion = "loo")
AI_VH_MR_A_m1 <- add_criterion(AI_VH_MR_A_m1, criterion = "loo")

loo_compare(VoiceHearing1,
           MindReading1,
           Apathy1,
           AI_VH_MR_m1,
           AI_VH_MR_A_m1)

## adding wights
loo_model_weights(VoiceHearing1,
           MindReading1,
           Apathy1,
           AI_VH_MR_m1,
           AI_VH_MR_A_m1)

```




## Second part

Q2.1) However, we know that the diagnosis is based on symptom assessment: if the overall sum of symptoms is severe enough, the participant gets a diagnosis. In other words, by selecting the patients, and including the symptoms in the model we might have inadvertently introduced an issue in our inference. Do try to draw a causal graph (Directed Acyclical Graph) of the variables and compare it with the types of causal graphs presented in the slides. Discuss which biases you might have introduced.

Q2.2.) Redesign your analysis following the graph and report how the results change
```{r}
# formula
champion_A_f <- bf(
  AltercentricIntrusion ~ 1 + Apathy)

# prior check
checkprior_champion <- check_prior(champion_A_f, d, priorR2,"AI~Apathy_allData")
checkprior_champion+ggtitle("AI~Apathy prior check using all data")


# model
champion_A_m <- brm(
    formula = champion_A_f,
    data = d,
    family = gaussian,
    prior = priorR2,
    sample_prior = T,
    refresh = 0,
    cores = cores,
    file = "Champion_Apathy_m"
    )

champion_A_m

#plot model

plot(champion_A_m)


# a plot for a posterior check
champion_posch <- pp_check(champion_A_m, nsamples = 100)
champion_posch+ggtitle("AI~Apathy Posterior Pedictive Check using all data")
  
# test and plot your hypothesis
champion_hyp <- hypothesis(champion_A_m, "Apathy < 0")
champion_hyp

plot(champion_hyp,plot = F)[[1]]+ggtitle("Using all data") 
plot(conditional_effects(champion_A_m), points = T,plot=F)[[1]]+ggtitle("Estimates of AI ~ Apathy using all data")

```

```{r, Theory correction}
# formula
TheoryFormula <- bf(
  AltercentricIntrusion ~ 1 + MindReading+VoiceHearing)

# prior check
checkprior_TheoryFormula <- check_prior(TheoryFormula, d, priorR2,"AI~VH_MR_AllData")
checkprior_TheoryFormula


# model
TheoryFormulaModel<- brm(
    formula = TheoryFormula,
    data = d,
    family = gaussian,
    prior = priorR2,
    sample_prior = T,
    refresh = 0,
    cores = cores,
    file = "TheoryModel_VH_MR"
    )

TheoryFormulaModel

#plot model

plot(TheoryFormulaModel)

# a plot for a posterior check
theory_posch <- pp_check(TheoryFormulaModel, nsamples = 100)
theory_posch +ggtitle("AI~MindReading+VoiceHearing Prior Check using all data")
  
# test and plot your hypothesis
theory_hyp_MR <- hypothesis(TheoryFormulaModel, "MindReading > 0")
theory_hyp_MR

theory_hyp_VH <- hypothesis(TheoryFormulaModel, "VoiceHearing > 0")
theory_hyp_VH

plot(theory_hyp_MR,plot=F)[[1]]+ggtitle("AI~MindReading+VoiceHearing Hypothesis Test using all data") 

plot(theory_hyp_VH,plot=F)[[1]]+ggtitle("AI~MindReading+VoiceHearing Hypothesis Test using all data")

plot(conditional_effects(TheoryFormulaModel), points = T)


```


```{r, Comparing better models}

## adding criterion loo
TheoryFormulaModel <-
  add_criterion(TheoryFormulaModel, criterion = "loo")
champion_A_m  <- add_criterion(champion_A_m, criterion = "loo")


loo_compare(TheoryFormulaModel,
            champion_A_m)

## adding wights
loo_model_weights(TheoryFormulaModel,
                  champion_A_m)

```
```{r, Building other models}

### voice hearing
VoiceHearing2 <- brm(
    formula = AI_VoiceHearing_f1,
    data = d,
    family = gaussian,
    prior = priorR2,
    sample_prior = T,
    refresh = 0,
    cores = cores,
    file = "VoiceHearing2"
    
  )
  VoiceHearing2
  
# a plot for a posterior check
VH_posch2 <- pp_check(VoiceHearing2, nsamples = 100)
VH_posch2
  
# test and plot your hypothesis
VH_hyp2 <- hypothesis(VoiceHearing2, "VoiceHearing > 0")
VH_hyp2

plot(VH_hyp2)  
plot(conditional_effects(VoiceHearing2), points = T)
  

## mind reading

MindReading2 <- brm(
    formula = AI_MindReading_f1,
    data = d,
    family = gaussian,
    prior = priorR2,
    sample_prior = T,
    refresh = 0,
    cores = cores,
    file = "MindReading2"
    )

MindReading2
# a plot for a posterior check
MR_posch2 <- pp_check(MindReading2, nsamples = 100)
MR_posch2
  
# test and plot your hypothesis
MR_hyp2 <- hypothesis(MindReading2, "MindReading > 0")
MR_hyp2

plot(MR_hyp2)  
plot(conditional_effects(MindReading2), points = T)
  
# VH + MR + A
AI_VH_MR_A_m2 <- brm(
  formula = AI_VH_MR_A_f1,
  data = d,
  family = gaussian,
  prior = priorR2,
  sample_prior = T,
  cores = cores,
  file = "AI_VH_MR_A_m2"
)

AI_VH_MR_A_m2

# Posterior predictive check
pp_check(AI_VH_MR_A_m2, nsamples = 100)

# Hypothesis testing + updating check

hypothes3_VH2 <- hypothesis(AI_VH_MR_A_m2,
           "VoiceHearing > 0")

hypothes3_MR2 <- hypothesis(AI_VH_MR_A_m2,
           "MindReading > 0")

hypothes3_Apathy2 <- hypothesis(AI_VH_MR_A_m2,
           "Apathy > 0")


hypothes3_VH2
hypothes3_MR2
hypothes3_Apathy2


plot(hypothes3_VH2)
plot(hypothes3_MR2)
plot(hypothes3_Apathy2)


## conditional effects
plot(conditional_effects(AI_VH_MR_A_m2), points = T)






```


```{r, Comparing better models2}

## adding criterion loo
VoiceHearing2 <- add_criterion(VoiceHearing2, criterion = "loo")
MindReading2 <- add_criterion(MindReading2, criterion = "loo")
champion_A_m  <- add_criterion(champion_A_m, criterion = "loo")
TheoryFormulaModel <- add_criterion(TheoryFormulaModel, criterion = "loo")
AI_VH_MR_A_m2 <- add_criterion(AI_VH_MR_A_m2, criterion = "loo")


loo_compare(VoiceHearing2,
            MindReading2,
            champion_A_m,
            TheoryFormulaModel,
            AI_VH_MR_A_m2)
## adding wights
loo_model_weights(VoiceHearing2,
            MindReading2,
            champion_A_m,
            TheoryFormulaModel,
            AI_VH_MR_A_m2)

```


## Third part

These issues are very difficult to think through, and not knowing the causal mechanisms generating the data in advance makes our inferences even more unreliable. To explore these issues, I recommend using simulations. In other words, defining a "true" model, generating data from it and assessing what different analyses would lead you to infer (and therefore which biases they might introduce). You can find the code I used to simulate your data below.

Q3.1) Look through the code and identify whether the results you have match the underlying truth. Discuss what you have learned.

Q3.2) OPTIONAL: is this a general pattern? Try varying the parameters (e.g. correlation values) and assess whether the new dataset(s) leads to the same biases in your analysis.



```{r}
pacman::p_load(MASS, tidyverse, psych)

seed <- 1981 # Defining a seed so the results are always the same
n <- 300 # Defining the amount of participants

SymptomCorr <- .2 # Defining the correlation of symptoms (as they tend to co-occur)
EffectCorrRel <- .2 # Defining the correlation between relevant symptoms and effect (Some symptoms are positively correlated with the effect)
EffectCorrIrrel <- 0 # Defining the correlation between irrelevant symptoms and effect (none)

# Creating the variance-covariance matrix for the variables we want to generate (3 symptoms, 1 effect)
Sigma <- matrix(data=c(1,SymptomCorr,SymptomCorr,EffectCorrRel,
                       SymptomCorr,1,SymptomCorr,EffectCorrRel,
                       SymptomCorr,SymptomCorr,1,EffectCorrIrrel,
                       EffectCorrRel,EffectCorrRel,EffectCorrIrrel,1),
                       nrow=4,ncol=4)

## Generate data from a multivariate (mvr) normal (n) distribution
data <- mvrnorm(n = n, # number of participant
        mu = c(1.2, 1.2, 1.2, 4), # mean of each variable
        Sigma) # variance co-variance matrix

# Giving meaningful names to variables and add ID
data <- data.frame(
  VoiceHearing = data[,1], 
  MindReading =  data[,2],
  Apathy =  data[,3], 
  AltercentricIntrusion = data[,4],
  ID = seq(nrow(data)))

# Assessing whether the participant has schizophrenia (high enough sum of symptoms)
# Here we choose participants scoring above 75% percentile (the most severe ones)
data$Diagnosis <- 0
data$Diagnosis[(data$VoiceHearing + data$MindReading + data$Apathy) > 
              quantile(data$VoiceHearing + data$MindReading + data$Apathy, .75)] <-1

## Plotting the relation between variables in schizophrenia
data1 <- data %>% subset(Diagnosis==1) %>% dplyr::select(-Diagnosis, -ID)
pairs.panels(data1)

## Plotting the relation between variables all participants
pairs.panels(dplyr::select(data,-Diagnosis, -ID))

write_csv(data, "datAss3.csv")
```

